Probabilistic forecasting of hourly Emergency Department arrivals

### 2022-11-17 paper/paper.Rmd

abstract: |
   An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients’ demand. It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals. We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.

@@

In this paper, we aim at filling several gaps, and our contributions to the literature are summarised as follows:

@@

5. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.

@@

Some studies employ models to forecast admissions and arrivals in Emergency Department. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss hourly ED forecasting.

@@

```{r tab-results, out.width= "50%", cache=FALSE}
results_table <- read_rds("results_table.rds") %>% arrange(`Quantile Bias`)
results_table <- results_table %>% 
  mutate(Method = 
           case_when(Method == "tbats" ~ "TBATS",
                     Method == "prophet" ~ "Prophet",
                     TRUE ~ Method))
knitr::kable(results_table, booktabs = T, 
             linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =9)
```

<!-- Why NAs for RMSE?? -->

In above figures, we only present the performance of the top seven methods. However, we have evaluated the performance of 14 methods. Evaluation metrics for all methods from the test period are presented in Table \@ref(tab:tab-results). They are ordered by Quantile Bias. The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.

@@

Finally, Figure \@ref(fig:time-accuracy) reports the forecast performance of each approach presented in Table \@ref(tab:tab-results) versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours. The X-axis shows the speed of each method presented as slow, moderate or fast. We observe that there is no clear association between speed and accuracy improvement of the models used in this study.

@@

Figure \@ref(fig:time-accuracy) shows that NOtr-2 is the fastest method that provides consistent accuracy assessed via all three accuracy measures. This figure also indicates that while the $t$ family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures. Therefore, if the speed is not a major constraint when generating the forecasts, one may employ this approach to generate forecasts.

>>>>

abstract: |
   An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients’ demand. It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals for a prediction window of 48 hours. We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.

@@

In this paper, we aim at filling these gaps and generate forecasts for a  prediction window of 48 hours. Our contributions to the literature are summarised as follows:

@@

5. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements for longer horizons and in other services [@al2021empirical], such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.

@@

Some studies employ models to forecast admissions and arrivals in Emergency Department [@gul2020exhaustive]. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss hourly ED forecasting.

@@

```{r tab-results, out.width= "50%", cache=FALSE}
results_table <- read_rds("results_table.rds") %>% arrange(`Quantile Bias`)
results_table <- results_table %>% mutate(Time=if_else(is.na(Time),1.279187,Time)) %>% rename(`Time (minutes)` = Time)
results_table <- results_table %>% 
  mutate(Method = 
           case_when(Method == "tbats" ~ "TBATS",
                     Method == "prophet" ~ "Prophet",
                     TRUE ~ Method))
knitr::kable(results_table, booktabs = T, 
             linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =9)
```

In above figures (Figures \@ref(fig:Pinball)- \@ref(fig:lead-time-rmse)), we only present the performance of the top seven methods. However, we have evaluated the performance of 14 methods from the test period. Evaluation metrics and computational time (in minutes) for all methods are presented in Table \@ref(tab:tab-results). They are ordered by Quantile Bias. The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.

@@

Finally, Figure \@ref(fig:time-accuracy) reports the forecast performance of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours, presented in Table \@ref(tab:tab-results). The X-axis shows the speed of each method presented as slow, moderate or fast. We observe that there is no clear association between speed and accuracy improvement of the models used in this study.

@@

Figure \@ref(fig:time-accuracy) shows that NOtr-2 is the fastest method that provides consistent accuracy assessed via all three accuracy measures. The figure also indicates that while the $t$ family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures. Therefore, if the speed is not a major constraint when generating the forecasts, one may employ this approach to generate forecasts.

### 2022-11-17 paper/paper.tex

  \begin{abstract}
  An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients' demand. It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals. We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
  \end{abstract}
   \begin{keyword} Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing\end{keyword}
 \end{frontmatter}

@@

In this paper, we aim at filling several gaps, and our contributions to the literature are summarised as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
\item
  We develop an advanced dynamic model to forecast ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
\item
  We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
\item
  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
\item
  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
\end{enumerate}

@@

Some studies employ models to forecast admissions and arrivals in Emergency Department. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss hourly ED forecasting.

@@

\caption{\label{tab:tab-results}Summary of studies in hourly emergency care forecasting}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{lrrrr}
\toprule
Method & Quantile Bias & Pinball & RMSE & Time\\
\midrule
NOtr-1 & 0.0098967 & 1.222583 & 0.2675957 & 451.6620471\\
ADAM-iETSX & 0.0104673 & 1.417260 & 0.0896228 & 92.9348605\\
NOtr-2 & 0.0118522 & 1.208561 & 0.2675957 & 86.5895462\\
Ttr-2 & 0.0140221 & 1.210108 & 0.3324146 & 956.5532849\\
Prophet & 0.0193799 & 1.447037 & 0.2955460 & 20.6755021\\
ETS & 0.0194389 & 1.434862 & 0.0121247 & 10.7175205\\
Poisson-1 & 0.0372137 & 1.204920 & 0.0095263 & 1.4763353\\
Poisson-2 & 0.0373884 & 1.188109 & 0.0082932 & 5.0768588\\
NBI-2 & 0.0540725 & 1.206241 & 0.3830272 & NA\\
Benchmark-2 & 0.0557392 & 1.217429 & 0.2592800 & 0.0947247\\
GBM-2 & 0.0600153 & 1.261690 & 1.7770897 & 602.4317496\\
TBATS & 0.0855702 & 1.536080 & 0.4859770 & 273.0558176\\
Regression-Poisson & 0.0929416 & 1.293524 & 0.8490258 & 67.1401641\\
Benchmark-1 & 0.1047874 & 1.254491 & 1.0042634 & 0.3874450\\
\bottomrule
\end{tabular}
\end{table}

In above figures, we only present the performance of the top seven methods. However, we have evaluated the performance of 14 methods. Evaluation metrics for all methods from the test period are presented in Table \ref{tab:tab-results}. They are ordered by Quantile Bias. The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.

@@

Finally, Figure \ref{fig:time-accuracy} reports the forecast performance of each approach presented in Table \ref{tab:tab-results} versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours. The X-axis shows the speed of each method presented as slow, moderate or fast. We observe that there is no clear association between speed and accuracy improvement of the models used in this study. We found that most of the methods considered in this paper fall into the slow category with a different range of performance, depending on the type of error measure. Figure \ref{fig:time-accuracy} also highlights that while NOtr-2 might be slow, but its accuracy is consistent when assessed using different accuracy measures. The figure also show that the \(t\) family GAMLSS model (Ttr-2) is the fastest model with a reasonably accurate performance across the three presented measures. Therefore, if the speed is a major constraint when generating the forecasts, one may employ this approach to generate forecasts.

>>>>

  \begin{abstract}
  An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients' demand. It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals for a prediction window of 48 hours. We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
  \end{abstract}
   \begin{keyword} Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing\end{keyword}
 \end{frontmatter}

@@

In this paper, we aim at filling these gaps and generate forecasts for a prediction window of 48 hours. Our contributions to the literature are summarised as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
\item
  We develop an advanced dynamic model to forecast ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
\item
  We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
\item
  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
\item
  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements for longer horizons and in other services \citep{al2021empirical}, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
\end{enumerate}

@@

Some studies employ models to forecast admissions and arrivals in Emergency Department \citep{gul2020exhaustive}. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss hourly ED forecasting.

@@

\caption{\label{tab:tab-results}Summary of studies in hourly emergency care forecasting}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{lrrrr}
\toprule
Method & Quantile Bias & Pinball & RMSE & Time (minutes)\\
\midrule
NOtr-1 & 0.0098967 & 1.222583 & 0.2675957 & 451.6620471\\
ADAM-iETSX & 0.0104673 & 1.417260 & 0.0896228 & 92.9348605\\
NOtr-2 & 0.0118522 & 1.208561 & 0.2675957 & 86.5895462\\
Ttr-2 & 0.0140221 & 1.210108 & 0.3324146 & 956.5532849\\
Prophet & 0.0193799 & 1.447037 & 0.2955460 & 20.6755021\\
ETS & 0.0194389 & 1.434862 & 0.0121247 & 10.7175205\\
Poisson-1 & 0.0372137 & 1.204920 & 0.0095263 & 1.4763353\\
Poisson-2 & 0.0373884 & 1.188109 & 0.0082932 & 5.0768588\\
NBI-2 & 0.0540725 & 1.206241 & 0.3830272 & 1.2791870\\
Benchmark-2 & 0.0557392 & 1.217429 & 0.2592800 & 0.0947247\\
GBM-2 & 0.0600153 & 1.261690 & 1.7770897 & 602.4317496\\
TBATS & 0.0855702 & 1.536080 & 0.4859770 & 273.0558176\\
Regression-Poisson & 0.0929416 & 1.293524 & 0.8490258 & 67.1401641\\
Benchmark-1 & 0.1047874 & 1.254491 & 1.0042634 & 0.3874450\\
\bottomrule
\end{tabular}
\end{table}

In above figures (Figures \ref{fig:Pinball}- \ref{fig:lead-time-rmse}), we only present the performance of the top seven methods. However, we have evaluated the performance of 14 methods from the test period. Evaluation metrics and computational time (in minutes) for all methods are presented in Table \ref{tab:tab-results}. They are ordered by Quantile Bias. The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecasts with Quantile Biases of 0.037 and above, ETS being the only exception with a value of 0.019.

@@

Finally, Figure \ref{fig:time-accuracy} reports the forecast performance of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours, presented in Table \ref{tab:tab-results}. The X-axis shows the speed of each method presented as slow, moderate or fast. We observe that there is no clear association between speed and accuracy improvement of the models used in this study.

We found that most of the methods considered in this paper fall into the fast category with a different range of performance, depending on the type of error measure. Among the fastest methods, ETS is very competitive when assessed using Quantile Bias and RMSE. Moreover, Poisson-1 and Poisson-2 provide accurate forecasts when evaluated by RMSE and Pinball.

Figure \ref{fig:time-accuracy} shows that NOtr-2 is the fastest method that provides consistent accuracy assessed via all three accuracy measures. The figure also indicates that while the \(t\) family GAMLSS model (Ttr-2) is the slowest method, it has a very good performance across the three presented accuracy measures. Therefore, if the speed is not a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
